State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-23 manuscript/manuscript_2/paper.tex
majority of Am comments

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. They modulate sinusoidally the times of arrival (TOAs) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple to search for a coincident GW signature. Congruent evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and a ``pulsar term" which are proportional to the GW strain at the Earth and the pulsar respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. It is acknowledged that, dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. \newline

\cite{KimpsonPTA} (K24 hereafter) introduced a state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, which self-consistently tracks the intrinsic, achromatic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of many standard PTA analyses and drops the pulsar term. In this paper we achieve two goals. First, we extend K24 to include the pulsar terms. We demonstrate how the static GW parameters and the GW phase at each pulsar can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Second, we quantify how including the pulsar term improves (i) the accuracy with which the static GW parameters are estimated, and (ii) the minimum detectable GW strain, compared to when the pulsar terms are omitted.  \newline

The paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in K24. In Section \ref{sec:pulsar_term} we show how to include the pulsar term via a convenient reparametrisation. In Section \ref{sec:rep_example1} we test the updated model, inclusive of the pulsar term, on synthetic data for a single representative GW source across an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:earth_vs_psr} we evaluate the parameter estimation accuracy and the minimum detectable GW strain. Conclusions are drawn in Section \ref{sec:discussion}. Throughout the paper we adopt natural units, with $c = G = \hbar = 1$, and the metric signature $(-,+,+,+)$. \newline

\section{State-Space Formulation}\label{sec:2}
In this section we briefly review the state-space formulation of a PTA experiment, as presented in K24. There are $N$ pulsars in the array, labelled $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation,  which describes secular braking combined with intrinsic, achromatic spin wandering (``timing noise''). The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation, which describes the TOA modulation induced by the GW. In Section \ref{sec:psr_frequency} we define and justify the phenomenological equation of motion for $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the static parameters of the model, including the GW source parameters, which are estimated ultimately from the data by nested sampling. The analysis method accepts as an input a sequence of pulse frequencies, instead of a sequence of pulse TOAs, in order to validate the approach in its simplest formulation and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs when implementing it to analyze real data, a subtle extension which we defer to a forthcoming paper.

\subsection{Spin evolution} \label{sec:psr_frequency}
We assume that the rest frame spin frequency of the $n$-th pulsar evolves according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ parametrizes the noise amplitude and produces characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking for the sake of definiteness, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars (MSPs), for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where $t_1$ labels the time of the first TOA. \newline

Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are not unique. Rather, they offer one possible phenomenological description consistent with the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. A phenomenological approach is obligatory, because a predictive, first-principles theory of timing noise does not currently exist, c.f. the multiple theorized mechanisms referenced in Section 1 of K24. Langevin equations like \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} have been applied successfully to analyse anomalous braking indices \citep{Vargas} and in hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. Idealisations include (i) the exclusion of elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the rest-frame spin frequency of the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, $f_{\rm p}$ is evaluated at the retarded time $t-d^{(n)}$, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
\end{align}
In Equation \eqref{eq:vareps}, $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is given by \citep[e.g.][]{Maggiore}
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ ,
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the GW amplitude tensor, and $\Phi_0$ is the phase offset of the GW with respect to some reference time. \newline

\noindent Equation \eqref{eq:g_func_trig} assumes that (i) the GW is a monochromatic plane wave; and (ii) $\boldsymbol{q}^{(n)}$ is constant. Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves by an amount \citep[e.g.][]{Sesana2010}
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \lesssim 20$ for $\Delta f_{\rm gw} \lesssim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsection{Static parameters}\label{sec:ss_params}
The model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, compromises $5N$ static parameters, that are specific to the pulsars in the array,
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
It also compromises seven parameters, that are specific to the GW source, viz. 
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
\end{equation}
where $h_0$ is the characteristic wave strain, $\iota$ is the orbital inclination, $\psi$ is the polarisation angle, $\delta$ is the declination and $\alpha$ is the right ascension. These parameters enter the model through Equation \eqref{eq:g_func_trig}, with $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$. The complete set of $7 + 5N$ static parameters is denoted by $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$. While we have no prior information about $\boldsymbol{\theta}_{\rm gw}$, there are constraints on $\boldsymbol{\theta}_{\rm psr}$ from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ typically \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source)

	 which also arise in traditional matched-filter analyses that exclude the pulsar terms.

 Including the pulsar terms also decreases the root-mean-square relative error in the estimated one-dimensional posteriors \textcolor{red}{by up to 99\%}, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quieter source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. They modulate sinusoidally the times of arrival (TOAs) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars to search for a coincident GW signature. Congruent evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs that are sufficiently close and massive may be resolvable individually but no definitive detection has been claimed so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

The modulation in a pulsar's TOAs due to a GW has two contributions: an ``Earth term" and ``pulsar terms" which are proportional to the GW strain at the Earth and every pulsar respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. It is acknowledged in standard analyses that dropping the pulsar terms leads to biases in the inferred parameters and reduces the detection probability, but both impacts are modest \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. \newline

\cite{KimpsonPTA} (K24 hereafter) introduced a state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, which self-consistently tracks the intrinsic, achromatic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 complements standard PTA analyses by tracking the specific, measured, time-ordered realization of the TOA fluctuations instead of fitting for their ensemble-averaged statistics (such as their power spectral density). The method described in K24 also follows the example of many standard PTA analyses and drops the pulsar terms. In this paper we achieve two goals. First, we extend K24 to include the pulsar terms. We demonstrate how the static GW parameters and the GW phase at each pulsar can be estimated successfully by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}, repeating the successful demonstration in K24 but with the pulsar terms now included. Second, we quantify how including the pulsar terms improves (i) the accuracy with which the static GW parameters are estimated, and (ii) the minimum detectable GW strain, compared to when the pulsar terms are omitted. We emphasize that the biases incurred by omitting the pulsar terms have been studied thoroughly in the context of standard PTA analyses \citep{Zhupulsarterms,Chen2022}; they are not new effects discovered here for the first time. Rather, the goal of this paper is to study them again in the context of the promising state-space formulation of the PTA analysis problem introduced by K24. The state-space formulation complements standard analysis techniques, e.g.\ based on matched filters \citep{PhysRevD.79.084030,PhysRevD.91.044048}. It does not supplant them and is likely to be most informative when run in tandem. \newline

The paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced by K24. In Section \ref{sec:pulsar_term} we show how to include the pulsar terms via a convenient reparametrisation. In Sections \ref{sec:rep_example1} and \ref{sec:pe_and_ms} we explain how to test the updated model, inclusive of the pulsar terms, by employing synthetic data for a single representative GW source across an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:earth_vs_psr} we quantify the parameter estimation accuracy and the minimum detectable GW strain, comparing the performance when the pulsar terms are included and excluded.. Conclusions are drawn in Section \ref{sec:discussion}. Throughout the paper we adopt natural units, with $c = G = \hbar = 1$, and the metric signature $(-,+,+,+)$. \newline

\section{State-Space Formulation}\label{sec:2}
In this section we briefly review the state-space formulation of a PTA experiment, as presented in K24. There are $N$ pulsars in the array, labelled $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation,  which describes secular braking combined with intrinsic, achromatic spin wandering (``timing noise''). The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation, which describes the TOA modulation induced by the GW. In Section \ref{sec:psr_frequency} we define and justify the phenomenological equation of motion for $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the static parameters of the model, including the GW source parameters, which are estimated ultimately from the data by nested sampling. The analysis method accepts as an input a sequence of pulse frequencies, instead of a sequence of pulse TOAs, in order to validate the approach in its simplest incarnation and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs when analyzing real data, a subtle extension which we defer to a forthcoming paper.

\subsection{Spin evolution} \label{sec:psr_frequency}
We assume that the rest frame spin frequency of the $n$-th pulsar evolves according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ parametrizes the noise amplitude and produces characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking for the sake of definiteness, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars (MSPs), for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where $t_1$ labels the first TOA. \newline

Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are not unique. Rather, they offer one possible phenomenological description consistent with the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. A phenomenological approach is obligatory, because a predictive, first-principles theory of timing noise does not currently exist, c.f. the multiple theorized mechanisms referenced in Section 1 of K24. Langevin equations like \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} have been applied successfully to analyse anomalous braking indices \citep{Vargas} and in hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, they are highly idealised \citep{Meyers2021,Myers2021MNRAS.502.3113M,2023MNRAS.520.2813A,Vargas}. Idealisations include (i) the exclusion of physics, which is likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the whitenes of $\xi^{(n)}(t)$ for MSPs in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$, noting that $\xi^{(n)}(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the rest-frame spin frequency of the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, $f_{\rm p}^{(n)}$ is evaluated at the retarded time $t-d^{(n)}$, and $\varepsilon^{(n)}(t)$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
\end{align}
In Equation \eqref{eq:vareps}, $\sigma_{\rm m}^2$ is the variance of the measurement noise at the telescope and is shared between all pulsars. In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is given by \citep[e.g.][]{Maggiore}
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ ,
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the GW amplitude tensor, and $\Phi_0$ is the phase offset of the GW with respect to some reference time. \newline

\noindent Equation \eqref{eq:g_func_trig} assumes that (i) the GW is a monochromatic plane wave; and (ii) $\boldsymbol{q}^{(n)}$ is constant. Regarding point (i), studies of SMBHB inspirals in the PTA context show that the GW frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves by an amount \citep[e.g.][]{Sesana2010}
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \lesssim 20$ for $\Delta f_{\rm gw} \lesssim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsection{Static parameters}\label{sec:ss_params}
The model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured} comprises $5N$ static parameters, that are specific to the pulsars in the array, viz.
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
It also comprises seven parameters, that are specific to the GW source, viz. 
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
\end{equation}
where $h_0$ is the characteristic wave strain, $\iota$ is the orbital inclination, $\psi$ is the polarisation angle, $\delta$ is the declination and $\alpha$ is the right ascension. These parameters enter the model through Equation \eqref{eq:g_func_trig}, with $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$. The complete set of $7 + 5N$ static parameters is denoted by $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$. While we have no prior information about $\boldsymbol{\theta}_{\rm gw}$, there are constraints on $\boldsymbol{\theta}_{\rm psr}$ from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ typically \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.
